\propertitle

### 2019-12-15 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-15T16:26:55+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The result is the distribution of the form
\begin{equation}
  \label{eq:app_maxent_pop_first}
  \begin{gathered}
  \px(\yA \| M, N)  = \frac{1}{Z(\yll)}\;
%  g(\yA)\;
  \exp\Biggl[\sum_{m}\yl_{m}\;
  \binom{\yA}{m}\bigg/\binom{N}{m}
  \Biggr]
  \\
\text{with}\qquad  Z(\yll)\defd \sum_{\yA}
  % g(\yA)\;
  \exp\Biggl[\sum_{m}\yl_{m}\;
  \binom{\yA}{m}\bigg/\binom{N}{m}
  \Biggr],
\end{gathered}
\end{equation}
where the $m'$ parameters $\yll \defd (\yl_{m})$ are determined by the
constraints~\eqref{eq:final_constraint}. This formula is further discussed
in appendix~\ref{sec:derivation_maxent_fullnet}.

The amount and degrees of the constraining moments depend on the questions
and hypotheses under study. We give some examples in the
next two sections.

Note that the constraint equation~\eqref{eq:final_constraint} for the
distribution~\eqref{eq:app_maxent_pop_first} may not have solutions
if $N$ is strictly larger than $n$, but it's still possible to find The possible discrepancy between its
two sides typically increases with the number of constraints. This
discrepancy comes from an approximation implicit in the maximum-entropy
method combined with one assumption specific to our application: that
measured averages equal expected values (or equivalently, the number of
time bins $T$ is infinite), and that an activity level $\yA$ can equally
likely be generated by any set of $\yA$ neurons in the larger population.
The magnitude of this discrepancy can be a signature of the presence of
neuronal \enquote{assemblies} or \enquote{clusters}
\citep[\chap~12]{gerstneretal2014}{hebb1949_r2002}. We discuss this matter
in \sect~\ref{sec:assumptions}.

@@

We apply the model just described to two data sets publicly available in
the literature:
\begin{itemize}[wide]
\item The \textbf{first data set}, from \textcite[rat
  14147]{stensolaetal2012}, consists of $n=65$ neurons from rat Medial
  Entorhinal Cortex, recorded for about 20 minutes. Their spikes are binned
  into $T=417\,641$ bins of $3\;\textrm{ms}$ width.
\item The \textbf{second data set}, from \textcite[data courtesy by A.
  Riehle and T. Brochier]{rostamietal2016_r2017}, %\texttt{data\_RS.npy}
  consists of $n=159$ neurons from macaque Motor Cortex, recorded for about
  15 minutes. Their spikes are binned into $T=300\,394$ bins of
  $3\;\textrm{ms}$ width.
\end{itemize}

@@

The second assumption concerns the latter distribution. Its density
function has one of the following expressions:
\begin{enumerate}[label=(\textit{\alph*}),wide]
\item an entropic prior \citep{skillingetal1984,rodriguez1991,neumann2007}
  \begin{subequations}\label{eq:maxent_prior}
    \begin{equation}
      \label{eq:entropic_prior}
      \pf(\ynu \| \yI) \propto  \exp[-\yL\;\sh(\yF;\yr)]
      \approx \binom{L}{L\nu_{0}, \dotsc, L\nu_{N}}\;
      \prod_{\yA}{r_{\yA}}^{L\nu_{\yA}}  
    \end{equation}
    where $\yr\defd(r_{\yA})$ is a reference distribution and the
    expression in larger parentheses is a multinomial coefficient;
  \item a \enquote{sufficiency prior}, from the exponential family, which
    follows from the Pitman-Koopman theorem if some moments $\set{m}$ are
    sufficient statistics for our belief about $\ynu$:
    \begin{equation}
      \label{eq:suff_prior}
      \pf(\ynu \| \yI) \propto
      \int\!\!\!\di\yla\;\pf(\yla \| \yI)\;
      \prod_{\yA}\delt\Biggl\{\ynu_{\yA}-
      \frac{r_{\yA}}{Z(\yla)}\,\exp\Bigl[
      \tsum_{m}\lambda_{m}\tbinom{\yA}{m}/\tbinom{N}{m}
      \Bigr]\Biggr\},
    \end{equation}
  \end{subequations}
  where $\yla\defd(\lambda_{m})$ are Lagrange multipliers and $Z(\yla)$ a
  normalization factor for the exponential distribution within the delta.
  This is a formal expression, the delta symbolizing that the possible
  domain of $\ynu$ is restricted from an $N$-dimensional simplex to a
  lower-dimensional subset thereof, parametrized by $\yla$.
\end{enumerate}

@@

It can be proved that the formulae above lead to the following belief
distribution for the joint frequency distribution $\yJ \defd (J_{\ya\yA})$
conditional on $\yf$:
\begin{multline}
  \label{eq:prob_joint_frequencies}
  \pf(\yJ \| \yf, \yI) \propto
                         \int\!\!\!\di\ynu\;\pf(\ynu\|\yI)\;
                         \binom{T}{TJ_{0 0},\dotsc,TJ_{n N}}\,
                         \prod_{\ya,\yA}\bigl(\yG_{\ya\yA}\;\nu_{\yA}\bigr)^{T\,J_{\ya\yA}}
  \approx{}  \\
\shoveright{\int\!\!\!\di\ynu\;\pf(\ynu\|\yI)\;
                         \exp\biggl(
-T \tsum_{\ya,\yA}J_{\ya\yA}\ln\frac{J_{\ya\yA}}{\yG_{\ya\yA}\;\nu_{\yA}}
                         \biggr)}
  \\
  \text{with}\quad \tsum_{\yA}J_{\ya\yA}=\yff_{\ya},
\end{multline}
where $\pf(\ynu\|\yI)$ is either of the
distributions~\eqref{eq:maxent_prior}. From this belief distribution we can
find the one for $\yFF_{\yA}\equiv \sum_{\ya}J_{\ya\yA}$ by
marginalization. Within the exponential we can recognize the relative
entropy of $\yJ$ with respect to $(\yG_{\ya\yA}\;\nu_{\yA})$, which is the
reason why these two joint distributions must be equal if $T$ tends to
infinity.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-15T17:33:33+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The result is the distribution of the form
\begin{equation}
  \label{eq:app_maxent_pop_first}
  \begin{gathered}
  \px(\yA \| M, N)  = \frac{1}{Z(\yll)}\;
%  g(\yA)\;
  \exp\Biggl[\sum_{m}\yl_{m}\;
  \binom{\yA}{m}\bigg/\binom{N}{m}
  \Biggr]
  \\
\text{with}\qquad  Z(\yll)\defd \sum_{\yA}
  % g(\yA)\;
  \exp\Biggl[\sum_{m}\yl_{m}\;
  \binom{\yA}{m}\bigg/\binom{N}{m}
  \Biggr],
\end{gathered}
\end{equation}
where the $m'$ parameters $\yll \defd (\yl_{m})$ are determined by the
constraints~\eqref{eq:final_constraint} and can be found by convex
extremization, discussed further in
appendix~\ref{sec:derivation_maxent_fullnet}. Note that for $N>n$ the
constraint equation~\eqref{eq:final_constraint} may never be exactly
satisfied by the distribution~\eqref{eq:app_maxent_pop_first} for any
$\yla$. In this case the extremization minimizes the discrepancy between
the left and right sides of~\eqref{eq:final_constraint}. This discrepancy
comes from an approximation implicit in the maximum-entropy method (namely
that the number of time bins $T$ is infinite) combined with one assumption
specific to our application: that an activity level $\yA$ can equally
likely be generated by any set of $\yA$ neurons in the larger population.
The magnitude of this discrepancy can be a signature of the presence of
neuronal \enquote{assemblies} or \enquote{clusters}
\citep[\chap~12]{gerstneretal2014}{hebb1949_r2002}. We discuss this matter
in \sect~\ref{sec:assumptions}.

@@

We apply the model just described to two data sets publicly available in
the literature:
\begin{itemize}[wide]
\item The \textbf{first data set}, from \textcite[rat
  14147]{stensolaetal2012}, consists of $n=65$ neurons from rat Medial
  Entorhinal Cortex, simultaneously recorded for about 20 minutes. Their
  spikes are binned into $T=417\,641$ bins of $3\;\textrm{ms}$ width.
\item The \textbf{second data set}, from \textcite[data courtesy by A.
  Riehle and T. Brochier]{rostamietal2016_r2017}, %\texttt{data\_RS.npy}
  consists of $n=159$ neurons from macaque Motor Cortex, simultaneously
  recorded for about 15 minutes. Their spikes are binned into $T=300\,394$
  bins of $3\;\textrm{ms}$ width.
\end{itemize}

@@

The second assumption concerns the latter distribution. Its density
function has one of the following expressions:
\begin{enumerate}[label=(\textit{\alph*}),wide]
\item an entropic prior \citep{skillingetal1984,rodriguez1991,neumann2007}
  \begin{subequations}\label{eq:maxent_prior}
    \begin{equation}
      \label{eq:entropic_prior}
      \pf(\ynu \| \yI) \propto  \exp[-\yL\;\sh(\yF;\yr)]
      \approx \binom{L}{L\nu_{0}, \dotsc, L\nu_{N}}\;
      \prod_{\yA}{r_{\yA}}^{L\nu_{\yA}}  
    \end{equation}
    where $\yr\defd(r_{\yA})$ is a reference distribution and the
    expression in larger parentheses is a multinomial coefficient;
  \item a \enquote{sufficiency prior}, from the exponential family, which
    follows from the Pitman-Koopman theorem if some moments $\set{m}$ are
    sufficient statistics for our belief about $\ynu$:
    \begin{equation}
      \label{eq:suff_prior}
      \pf(\ynu \| \yI) \propto
      \int\!\!\!\di\yla\;\pf(\yla \| \yI)\;
      \prod_{\yA}\delt\Biggl\{\ynu_{\yA}-
      \frac{r_{\yA}}{Z(\yla)}\,\exp\Bigl[
      \tsum_{m}\lambda_{m}\tbinom{\yA}{m}/\tbinom{N}{m}
      \Bigr]\Biggr\},
    \end{equation}
  \end{subequations}
  where $\yla\defd(\lambda_{m})$ are Lagrange multipliers and $Z(\yla)$ a
  normalization factor for the exponential distribution within the delta.
  This is a formal expression, the delta symbolizing that the possible
  domain of $\ynu$ is restricted from an $N$-dimensional simplex to a
  lower-dimensional subset thereof, parametrized by $\yla$. When this prior
  is used, all integrations over $\ynu$ are replaced by integrations over
  $\yla$.
\end{enumerate}

@@

It can be proved that the formulae above lead to the following belief
distribution for the joint frequency distribution $\yJ \defd (J_{\ya\yA})$
and the long-run frequency distribution $\ynu$, conditional on $\yf$:
\begin{multline}
  \label{eq:prob_joint_frequencies}
  \pf(\yJ, \ynu \| \yf, \yI) \propto
  %\int\!\!\!\di\ynu\;
  \pf(\ynu\|\yI)\;
                         \binom{T}{TJ_{0 0},\dotsc,TJ_{n N}}\,
                         \prod_{\ya,\yA}\bigl(\yG_{\ya\yA}\;\nu_{\yA}\bigr)^{T\,J_{\ya\yA}}
  \approx{}  \\
  \shoveright{%\int\!\!\!\di\ynu\;
    \pf(\ynu\|\yI)\;
                         \exp\biggl(
-T \tsum_{\ya,\yA}J_{\ya\yA}\ln\frac{J_{\ya\yA}}{\yG_{\ya\yA}\;\nu_{\yA}}
                         \biggr)}
  \\
  \text{with}\quad \tsum_{\yA}J_{\ya\yA}=\yff_{\ya},
\end{multline}
where $\pf(\ynu\|\yI)$ is either of the
distributions~\eqref{eq:maxent_prior}. Note the relative entropy in the
exponential. If we want to conditionalize not on $\yf$ but on a set of
measured averages $C(m) = \sum_{\ya}c_{\ya}(m) \yff_{\ya}$, $m\in M$, such
as~\eqref{eq:chain_prescription_moments}, we have to replace the
integration constraint $\sum_{\yA}J_{\ya\yA}=\yff_{\ya}$ with
$\sum_{\ya\yA}c_{\ya}(m)\,J_{\ya\yA}=C(m)$.

From this belief distribution we can find the one for
$\yFF_{\yA}\equiv \sum_{\ya}J_{\ya\yA}$ and any others of interest by
marginalization. Within the exponential we can recognize the relative
entropy of $\yJ$ with respect to $(\yG_{\ya\yA}\;\nu_{\yA})$, which is the
reason why these two joint distributions must be equal if $T$ tends to
infinity.
